Numerical Stability of Detonations in White Dwarf Simulations

### 2019-02-06 papers/ignition/paper.tex
Mention density at the ignition time

The simulation is terminated when the peak temperature on the domain first reaches
$4 \times 10^9$ K, which we call a thermonuclear ignition. This stopping criterion is a
proxy for the beginning of a detonation. Reaching this temperature does not guarantee
that a detonation will begin, and in this study we do not directly address the question
of whether a ignition of this kind always leads to a detonation. \added{Nor are we commenting
on the physics of the ignition process itself.} Rather, the main question
we investigate here is whether this ignition is numerically converged, \added{and for this purpose
this arbitrary stopping point is sufficient, since in a converged simulation the stopping point
should be reached at the same time independent of resolution}. A converged ignition
is a prerequisite to having a converged detonation. We measure two diagnostic quantities:
the time since the beginning of the simulation required to reach this ignition criterion,
and the distance from the contact point of the peak temperature.

@@

We find that for our test problem (and also the WD collisions we have simulated)
we do observe $\tau_s > \tau_e$; typically the ratio is a factor of 2--5 at
low resolution. This means that an ignition is very likely to occur for numerical
reasons, regardless of whether it would occur for physical reasons. \added{At low
resolution, adding more resolution does not meaningfully improve the ratio of
$\tau_s$ to $\tau_e$ at the point of ignition. The ignition timescale is so short
that virtually all of the energy release occurs in a single timestep even if the
timestep gets shorter due to the CFL limiter. It is only when the resolution gets
sufficiently high that we can simultaneously resolve the energy release over multiple
timesteps and the advection of energy across multiple zones. Even at the highest resolution
we could achieve, about 50 cm, $\tau_s / \tau_e$ was 0.8 at ignition, which is not
sufficiently small to be confident of numerical stability at that resolution. Note
that merely decreasing the timestep (at fixed resolution) does not help here, as the
instability criterion is, to first order, independent of the size of the timestep.}

>>>>

The simulation is terminated when the peak temperature on the domain first reaches
$4 \times 10^9$ K, which we call a thermonuclear ignition (\added{for reference, the
density at the location where the ignition occurs is approximately $1.4\times 10^7\ \text{g / cm}^3$}). This stopping criterion is a
proxy for the beginning of a detonation. Reaching this temperature does not guarantee
that a detonation will begin, and in this study we do not directly address the question
of whether a ignition of this kind always leads to a detonation. \added{Nor are we commenting
on the physics of the ignition process itself.} Rather, the main question
we investigate here is whether this ignition is numerically converged, \added{and for this purpose
this arbitrary stopping point is sufficient, since in a converged simulation the stopping point
should be reached at the same time independent of resolution}. A converged ignition
is a prerequisite to having a converged detonation. We measure two diagnostic quantities:
the time since the beginning of the simulation required to reach this ignition criterion,
and the distance from the contact point of the peak temperature.

@@

We find that for our test problem (and also the WD collisions we have simulated)
we do observe $\tau_s > \tau_e$; typically the ratio is a factor of 2--5 at
low resolution. This means that an ignition is very likely to occur for numerical
reasons, regardless of whether it would occur for physical reasons. \added{At low
resolution, adding more resolution does not meaningfully improve the ratio of
$\tau_s$ to $\tau_e$ at the point of ignition. The ignition timescale is so short
that virtually all of the energy release occurs in a single timestep even if the
timestep gets shorter due to the CFL limiter. It is only when the resolution gets
sufficiently high that we can simultaneously resolve the energy release over multiple
timesteps and the advection of energy across multiple zones. Even at the highest resolution
we could achieve, about 50 cm, $\tau_s / \tau_e$ was 0.8 at ignition, which is not
sufficiently small to be confident of numerical stability at that resolution. Note
that merely decreasing the timestep (at fixed resolution) does not help here either, as the
instability criterion is, to first order, independent of the size of the timestep.}
